The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles

### 2022-11-09 ms/ms.tex
<edit ms>

As noted already, this prevalence was remarked on almost 50 years
earlier, the result of large-scale collections acquisitions that major
museums had engaged in since the late 1800s (e.g., Květinová
2011:65-66). But Chimú pottery continued to flow into established museum
collections, and to be sought as new museums were established (see Mowat
1988 for an example of donation and accessioning practices).
Collection-building accelerated as new museums were founded after World
War II and continued with limited constraints until after 1970, when the
UNESCO Convention regarding cultural property established guidelines
that gradually altered legal standards and ethical practices surrounding
collecting (Gerstenblith 2013). Although some museums continued to send
curators to Peru to purchase from antiquities dealers, a cursory review
of museum records reveals that many new accessions came as gifts from
individual donors, who provided little or no provenance information. For
example, the Metropolitan Museum of Art received a feline bottle
(64.228.17) from Nathan Cummings in 1964, who had purchased the vessel a
decade earlier from a collector in Buenos Aires. The piece was one of
more than 1000 artifacts that Cummings is credited with donating to the
museum. The Cleveland Museum of Art, founded in 1913, also built its
collections through donations, including a blackware jar (1959.333)
given by William Ellery Greene in 1959. It was the only Andean artifact
given to the museum by the donor.

Building on preliminary work by Pedro Rojas Ponce and Dorothy Menzel,
Sheele and Patterson (1966) compared illustrations from the published
literature with vessels from Peru's Museo Nacional de Antropología y
Arqueología and Harvard University's Peabody Museum of Archaeology and
Ethnology. They created a preliminary seriation of Chimú ceramics from
the end of the Middle Horizon to the early Colonial Period,
distinguishing seven different phases based on the identification of
different vessel shapes, decorative elements, and firing
characteristics. Although the seriation was developed to build a more
fine-grained relative chronology, the small sample sizes of some phases
presented interpretive challenges. For example, the fourth phase
\emph{Lambayeque}, was based on seven vessels from a single grave in the
Lambayeque Valley, located approximately 175 km to the north of Chan
Chan, and probably reflects regional ceramic diversity more than a
distinct period of Chimú production in the Moche Valley. Other phases
were based on small sample sizes and vessels with limited provenance
information. One important distinction in the Scheele and Patterson
seriation was between Chimú ceramics of the late LIP (Chimu Phase T-1)
and those from the time of the Inca occupation (Chimu Phase T-2). The
former was defined based on the prevalence of stirrup-spout vessels and
the \emph{less carefully made} production of mold-produced vessels that
seemed to represent a more restricted set of shapes than the preceding
phase (Trujillo Phase T-2). The Chimú-Inca phase maintained elements of
existing ceramic production, adding features from Inca vessel shapes.
Scheele and Patterson note (1966:24) that this phase seems to be
distributed beyond earlier Chimu political boundaries and economic
networks---it has been identified on the south and central coast and in
the Cuzco region. Overall, Scheele and Patterson's work clarifies how
some vessel shapes developed over time, making it possible to assign
provenienced gravelots to more specific late prehispanic periods. It
also identified some important aspects of production practices and
hybridity that have been important for subsequent archaeological
fieldwork.

@@

Although the first attempts at defining Chimú ceramics within a broader
north coast sequence utilized museum collections, archaeological
excavations have increasingly contributed to the reconstruction of
ceramic production practices. In the mid-20th century, excavators turned
to cemeteries as a source of well-provenienced vessels that could be
used to elaborate the regional sequence (e.g., Collier 1955; Willey
1947). Such work represented only a miniscule proportion of Chimú
ceramics that were being dug up at that time, most of which were
obtained by huaqueros whose illicit work supplied private collectors and
museums around the world. Studies of whole vessels concluded that the
use of molds increased in late prehispanic periods on the north coast
(Collier 1955), and by the 1960s, some researchers had identified molds
and matrices in the pottery collections acquired by some museums (e.g.,
Grossman 1969-1970; Thompson 1963). Such work constituted an important
step toward moving from straightforward aesthetic considerations to
questions about the production and consumption of Chimú ceramics, but it
remained rooted in mortuary collections that did not focus on evidence
of production practices.

At the end of the 1960s, Moseley and Mackey's Chan Chan-Moche Valley
Project (1969-1975) began to increase the intensity of professional
archaeological research, with regional survey work in the lower Moche
Valley and mapping and excavations at Chan Chan and other nearby sites.
While not focused on cemetery sites, this project and related studies
encountered additional mortuary contexts with well-preserved ceramics
(see Donnan and Mackey 1978). It was not until the 1990s, however, that
archaeologists began to identify loci of ceramic production based on the
presence of overfired sherds, molds and matrices, stamps, tools, kilns,
and other evidence (Donnan 1997; Hayashida 1999; Mackey 2003; Mackey and
Sapp 2021; Tschauner 2006, 2009). Some of the workshops reflect ceramic
production practices during the preceding era of Chimú imperial
expansion, whereas others were in use during the Inca occupation of the
north coast. Surface collections and excavations in production areas
generated multiple lines of material evidence, including large numbers
of potsherds that could not easily be assigned to the diverse array of
shapes identified in collection-based classifications.

The study of Chimú and Chimú-Inca pottery workshops has focused on two
central themes: specialization and cultural (dis)continuity under
foreign rule. These issues were nascent in many of the earlier
collection-based studies, but the careful excavation and analysis of
production contexts added invaluable evidence to the discussion. Molds
and matrices encountered at workshops indicate that multiple vessel
parts were mass-produced and had to be assembled following a specific
sequence, which Tschauner (2006:179) interprets as evidence of
specialized production during Chimú times. Tschauner's excavations at
the Pampa de Burros workshop in the Lambayeque Valley identified heavy
use of vertical half-molds to produce jars, canteens or flasks, bottles,
and ollas---virtually all of the shapes that were used by Chimú-era
settlements in the surrounding valley (2006:182). He sees the local
potters as relatively unskilled producers who relied on more adept
makers of ceramic molds, and whose \emph{market-oriented} work could
flexibly adapt to consumer preferences (2006:183-185). Levine (2011)
builds on this idea, noting that some of the variability seen in the
Chimú assemblage could come from ceramic producers \emph{mixing and
matching} molds for different vessel parts and adornments to produce new
combinations.

For the Inca period, scholars acknowledge a degree of continuity in
local potting practices, but also some significant changes. At Tambo
Real and La Viña in the Leche Valley, Hayashida (1999) notes the
continuity of local practices for shaping vessels (press molds, paddle
and anvil) and firing. While local jar shapes persisted in Inca times,
she observes the presence of a mold for an Inca flared jar neck
(1999:345), which could be added to jars that were shaped using local
press molds. Overall, Hayashida sees little evidence of Inca
\emph{retraining} of Chimú potters.

Other Inca-Chimú workshops in the Jequetepeque Valley provide evidence
that practices varied at different production locations. Donnan's
(1997:32-35) surface collections at Cañoncillo encountered dozens of
mold fragments, most of which were for shaping the bottom and top
hemispheres of ollas. Donnan identified molds for shaping jars along a
vertical axis, and for producing bottles, including the distinctive Inca
``aryballoid'' shape. Although he reports that no molds appeared for
shaping handles or rims, there were molds for producing stirrup spouts,
as well as for humans, birds, and other animals. The production evidence
from this site suggests that while some vessels were largely shaped
using molds, some elements were hand-formed and showed variations
suggesting the work of multiple potters. At the Inca administrative
center of Farfán, Mackey (2004:336-338) excavated a small workshop that
produced large jars (tinajas) for brewing or storage, using the
paddle-and-anvil technique. Finally, the excavation of three workshop
patios at El Algarrobal de Moro, a lower-order Chimú administrative
center, encountered 25 mold fragments, including molds for a stirrup
spout vessel and a miniature jar (Mackey and Sapp 2021:21-22).

@@

For our broader study, we scanned \#\#\#\# prehispanic ceramic vessels
from two collections that reflect the broader history of collecting and
curating Chimú ceramics. Our first collection was the Art and Art
History Collection (AAHC), a modest sized collection that accumulated
haphazardly at the University of Texas at Austin as numerous individuals
donors gave the university artifacts that had little or no provenance
information. The ancient pottery now in the AAHC was previously curated
by the Texas Memorial Museum, which transferred them to the department
of Art and Art History in 2004 as the museum implemented new curatorial
priorities. In 2018, we scanned 112 Andean vessels from the AAHC
collection. The data presented here come from a sample of Chimú bottles
(n=23) and jars (n=16) from the UT collection. \emph{\textbf{{[}Astrid:
How many Andean vessels overall in the collection, and how much Chimú?
Should we include a table of the samples here, with as much provenance
info as we can find?{]}}}

As a second collection, we also worked with a large museum sample that
was acquired in Peru by a professional archaeologist the early days of
museum collecting. The Bandelier Collection at the American Museum of
Natural History consists of nearly 8,000 Andean artifacts from Peru and
Bolivia. The Swiss-American archaeologist Adolph Bandelier acquired
these pieces as part of his 1890s Andean expedition, which was initially
funded by Henry Villard. Bandelier visited the Trujillo area in 1893,
where he sketched the ruins of Chan Chan and acquired more than 400
Chimú vessels and other artifacts that were accessioned by the museum.
Although the records of Bandelier's expedition remain unpublished, the
collection offers a large sample size of vessels that were collected at
the same time, during a period of archaeological collecting when dealers
were less attuned to the targeted collecting that Kroeber observed a
generation later. In 2020, we scanned \#\#\# Chimú, Inca, and Chimú-Inca
ceramic vessels at the AMNH.

The exploratory nature of the work also raised ethical considerations
regarding work with museum collections. Archaeologists have intensified
their work with museum collections in recent years, and the
consideration of whether to conduct such research are as important as
the logistical questions of how to accomplish the work. The ethical
principles of the Society for American Archaeology highlight the values
of stewardship and public responsibility---it is important for
archaeologists to care for the material record, and also to share
information about it with the public---identifying ``public education
and outreach'' as significant. From this perspective, the
non-destructive and museum-based work of 3D scanning presented little
risk to the artifacts being analyzed, and our research was designed to
be available to the public. When working with the AAHC vessels, we
developed an educational component, including a small class of UT
undergraduates in the lab-based collection of linear measurements of the
vessels being scanned. That supervised research provided a diverse group
of undergraduates with artifact analysis, while also teaching them about
the loss of heritage and scientific knowledge that accompanies the
collection of antiquities.

While the careful study and public reporting of museum collections is
consistent with professional ethics in Americanist archaeology, the SAA
also identifies a potential ethical concern in doing so: the
commercialization of the archaeological record. The SAA clearly states
that ``Archaeologists should therefore carefully weigh the benefits to
scholarship of a project against the costs of potentially enhancing the
commercial value of archaeological objects. Whenever possible they
should discourage, and should themselves avoid, activities that enhance
the commercial value of archaeological objects, especially objects that
are not curated in public institutions, or readily available for
scientific study, public interpretation, and display.'' As described
above, Chimú blackware pottery is an artifact class that was heavily
collected in the past, and it is openly sold today by galleries and
internet vendors. As Atwood (2004:22-23) notes, huaqueros, dealers, and
collectors are no longer interested in trafficking Andean ceramics,
especially mass-produced wares like those of the Chimú Empire. Our work
with publicly-held collections poses little risk of increasing the
present commercial value of Chimú ceramics, while its public discussion
of the losses that accompany undocumented excavations and illicit
collection constitutes a benefit that outweighs any associated risks.

@@

Linear shape variables\ldots{} (Figure \ref{fig:linear}).

@@

Landmark data were aligned to a global coordinate system
\citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through
generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R
using the \texttt{geomorph} and \texttt{RRPP} packages
\citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition
translates and rotates the coordinate data to allow for comparisons
among objects, while also scaling each biface using unit-centroid
size---the square root of the sum of squared distances from each
landmark to the specimen's centroid
\citep{RN11139,RN11140,RN11564,RN478}. The \texttt{geomorph} package
uses a partial Procrustes superimposition that projects the aligned
specimens into tangent space subsequent to alignment in preparation for
the use of multivariate methods that assume linear space (Figure
\ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

@@

Linear measurements from the sample of jars and bottles included vessel
height, rim thickness, and the diameter of each vessel at its orifice,
neck, belly, and base. Of these metrics, vessel height, base diameter,
and neck diameter did not differ significantly between the two samples.
Overall, the jars and bottles in our pilot sample overlap in some key
dimensions, with the bottles showing greater overall size variations,
which would be not be consistent with production using a single set of
molds. The other three sets of linear measurements showed significant
differences between the jars and bottles in our sample, in terms of
orifice diameter, belly diameter, and rim thickness. Given the different
production scenarios emerging from archaeological field analyses---a
combination of hand-shaping and use of whole vessel or elemental molds,
the significant differences in orifice and belly diameters might
indicate that the kinds of bottles and jars in our sample were produced
as distinct shapes, rather than by mixing and matching different molds
to cater to changing consumer tastes. The differences in rim thickness
suggest that distinct production practices were used when producing
bottles and jars. {[}\emph{\textbf{Zac: Should we discuss the analysis
of the figures on the vessels, and then point out that it's
inconclusive, or just leave that out?}}{]}

>>>>

As noted already, this prevalence was remarked on almost 50 years
earlier, the result of large-scale collections acquisitions that major
museums had engaged in since the late 1800s \citep[65-66]{RN11159}. But
Chimú pottery continued to flow into established museum collections, and
to be sought as new museums were established (see Mowat 1988 for an
example of donation and accessioning practices). Collection-building
accelerated as new museums were founded after World War II and continued
with limited constraints until after 1970, when the UNESCO Convention
regarding cultural property established guidelines that gradually
altered legal standards and ethical practices surrounding collecting
\citep{RN11160}. Although some museums continued to send curators to
Peru to purchase from antiquities dealers, a cursory review of museum
records reveals that many new accessions came as gifts from individual
donors, who provided little or no provenance information. For example,
the Metropolitan Museum of Art received a feline bottle (64.228.17) from
Nathan Cummings in 1964, who had purchased the vessel a decade earlier
from a collector in Buenos Aires. The piece was one of more than 1000
artifacts that Cummings is credited with donating to the museum. The
Cleveland Museum of Art, founded in 1913, also built its collections
through donations, including a blackware jar (1959.333) given by William
Ellery Greene in 1959. It was the only Andean artifact given to the
museum by the donor.

Building on preliminary work by Pedro Rojas Ponce and Dorothy Menzel,
\citet{RN11158} compared illustrations from the published literature
with vessels from Peru's Museo Nacional de Antropología y Arqueología
and Harvard University's Peabody Museum of Archaeology and Ethnology.
They created a preliminary seriation of Chimú ceramics from the end of
the Middle Horizon to the early Colonial Period, distinguishing seven
different phases based on the identification of different vessel shapes,
decorative elements, and firing characteristics. Although the seriation
was developed to build a more fine-grained relative chronology, the
small sample sizes of some phases presented interpretive challenges. For
example, the fourth phase \emph{Lambayeque}, was based on seven vessels
from a single grave in the Lambayeque Valley, located approximately 175
km to the north of Chan Chan, and probably reflects regional ceramic
diversity more than a distinct period of Chimú production in the Moche
Valley. Other phases were based on small sample sizes and vessels with
limited provenance information. One important distinction in the Scheele
and Patterson seriation was between Chimú ceramics of the late LIP
(Chimu Phase T-1) and those from the time of the Inca occupation (Chimu
Phase T-2). The former was defined based on the prevalence of
stirrup-spout vessels and the \emph{less carefully made} production of
mold-produced vessels that seemed to represent a more restricted set of
shapes than the preceding phase (Trujillo Phase T-2). The Chimú-Inca
phase maintained elements of existing ceramic production, adding
features from Inca vessel shapes. \citet[24]{RN11158} note that this
phase seems to be distributed beyond earlier Chimu political boundaries
and economic networks---it has been identified on the south and central
coast and in the Cuzco region. Overall, Scheele and Patterson's work
clarifies how some vessel shapes developed over time, making it possible
to assign provenienced gravelots to more specific late prehispanic
periods. It also identified some important aspects of production
practices and hybridity that have been important for subsequent
archaeological fieldwork.

@@

Although the first attempts at defining Chimú ceramics within a broader
north coast sequence utilized museum collections, archaeological
excavations have increasingly contributed to the reconstruction of
ceramic production practices. In the mid-20th century, excavators turned
to cemeteries as a source of well-provenienced vessels that could be
used to elaborate the regional sequence \citep{RN8727,RN11161}. Such
work represented only a miniscule proportion of Chimú ceramics that were
being dug up at that time, most of which were obtained by huaqueros
whose illicit work supplied private collectors and museums around the
world. Studies of whole vessels concluded that the use of molds
increased in late prehispanic periods on the north coast \citep{RN8727},
and by the 1960s, some researchers had identified molds and matrices in
the pottery collections acquired by some museums
\citep{RN11162,RN11163}. Such work constituted an important step toward
moving from straightforward aesthetic considerations to questions about
the production and consumption of Chimú ceramics, but it remained rooted
in mortuary collections that did not focus on evidence of production
practices.

At the end of the 1960s, Moseley and Mackey's Chan Chan-Moche Valley
Project (1969-1975) began to increase the intensity of professional
archaeological research, with regional survey work in the lower Moche
Valley and mapping and excavations at Chan Chan and other nearby sites.
While not focused on cemetery sites, this project and related studies
encountered additional mortuary contexts with well-preserved ceramics
\citep{RN11165}. It was not until the 1990s, however, that
archaeologists began to identify loci of ceramic production based on the
presence of overfired sherds, molds and matrices, stamps, tools, kilns,
and other evidence
\citep{RN11164,RN8875,RN11166,RN11167,RN11168,RN11169}. Some of the
workshops reflect ceramic production practices during the preceding era
of Chimú imperial expansion, whereas others were in use during the Inca
occupation of the north coast. Surface collections and excavations in
production areas generated multiple lines of material evidence,
including large numbers of potsherds that could not easily be assigned
to the diverse array of shapes identified in collection-based
classifications.

The study of Chimú and Chimú-Inca pottery workshops has focused on two
central themes: specialization and cultural (dis)continuity under
foreign rule. These issues were nascent in many of the earlier
collection-based studies, but the careful excavation and analysis of
production contexts added invaluable evidence to the discussion. Molds
and matrices encountered at workshops indicate that multiple vessel
parts were mass-produced and had to be assembled following a specific
sequence, which \citet[179]{RN11168} interprets as evidence of
specialized production during Chimú times. Tschauner's excavations at
the Pampa de Burros workshop in the Lambayeque Valley identified heavy
use of vertical half-molds to produce jars, canteens or flasks, bottles,
and ollas---virtually all of the shapes that were used by Chimú-era
settlements in the surrounding valley \citep[182]{RN11168}. He sees the
local potters as relatively unskilled producers who relied on more adept
makers of ceramic molds, and whose \emph{market-oriented} work could
flexibly adapt to consumer preferences \citep[183-185]{RN11168}.
\citet{RN11170} builds on this idea, noting that some of the variability
seen in the Chimú assemblage could come from ceramic producers
\emph{mixing and matching} molds for different vessel parts and
adornments to produce new combinations.

For the Inca period, scholars acknowledge a degree of continuity in
local potting practices, but also some significant changes. At Tambo
Real and La Viña in the Leche Valley, \citet{RN8875} notes the
continuity of local practices for shaping vessels (press molds, paddle
and anvil) and firing. While local jar shapes persisted in Inca times,
she observes the presence of a mold for an Inca flared jar neck
\citep[345]{RN8875}, which could be added to jars that were shaped using
local press molds. Overall, Hayashida sees little evidence of Inca
\emph{retraining} of Chimú potters.

Other Inca-Chimú workshops in the Jequetepeque Valley provide evidence
that practices varied at different production locations. The surface
collections of \citet[32-35]{RN11164} at Cañoncillo encountered dozens
of mold fragments, most of which were for shaping the bottom and top
hemispheres of ollas. Donnan identified molds for shaping jars along a
vertical axis, and for producing bottles, including the distinctive Inca
\emph{aryballoid} shape. Although he reports that no molds appeared for
shaping handles or rims, there were molds for producing stirrup spouts,
as well as for humans, birds, and other animals. The production evidence
from this site suggests that while some vessels were largely shaped
using molds, some elements were hand-formed and showed variations
suggesting the work of multiple potters. At the Inca administrative
center of Farfán, \citet[336-338]{RN11166} excavated a small workshop
that produced large jars (\emph{tinajas}) for brewing or storage, using
the paddle-and-anvil technique. Finally, the excavation of three
workshop patios at El Algarrobal de Moro, a lower-order Chimú
administrative center, encountered 25 mold fragments, including molds
for a stirrup spout vessel and a miniature jar \citep[21-22]{RN11167}.

@@

For our broader study, we scanned \#\#\#\# prehispanic ceramic vessels
from two collections that reflect the broader history of collecting and
curating Chimú ceramics. Our first collection was the Art and Art
History Collection (AAHC), a modest sized collection that accumulated
haphazardly at the University of Texas at Austin as numerous individuals
donors gave the university artifacts that had little or no provenance
information. The ancient pottery now in the AAHC was previously curated
by the Texas Memorial Museum, which transferred them to the department
of Art and Art History in 2004 as the museum implemented new curatorial
priorities. Those data presented here come from a sample of Chimú
bottles (n=23) and jars (n=16) from the UT collection.
\emph{\textbf{{[}Astrid: How many Andean vessels overall in the
collection, and how much Chimú? Should we include a table of the samples
here, with as much provenance info as we can find?{]}}}

The exploratory nature of this effort also raised ethical considerations
regarding modern research with extant museum collections. Archaeologists
have intensified their work with museum collections in recent years, and
the consideration of whether to conduct such research are as important
as the logistical questions of how to accomplish the work. The ethical
principles of the Society for American Archaeology highlight the values
of stewardship and public responsibility---it is important for
archaeologists to care for the material record, and also to share
information about it with the public---identifying ``public education
and outreach'' as significant. From this perspective, the
non-destructive and museum-based work of 3D scanning presented little
risk to the artifacts being analyzed, and our research was designed to
be available to the public. When working with the AAHC vessels, we
developed an educational component, including a small class of UT
undergraduates in the lab-based collection of linear measurements of the
vessels being scanned. That supervised research provided a diverse group
of undergraduates with artifact analysis, while also teaching them about
the loss of heritage and scientific knowledge that accompanies the
collection of antiquities.

While the careful study and public reporting of museum collections is
consistent with professional ethics in Americanist archaeology, the SAA
also identifies a potential ethical concern in doing so: the
commercialization of the archaeological record. The SAA clearly states
that ``Archaeologists should therefore carefully weigh the benefits to
scholarship of a project against the costs of potentially enhancing the
commercial value of archaeological objects. Whenever possible they
should discourage, and should themselves avoid, activities that enhance
the commercial value of archaeological objects, especially objects that
are not curated in public institutions, or readily available for
scientific study, public interpretation, and display.'' As described
above, Chimú blackware pottery is an artifact class that was heavily
collected in the past, and it is openly sold today by galleries and
internet vendors. As \citet[22-23]{RN11153} notes, huaqueros, dealers,
and collectors are no longer interested in trafficking Andean ceramics,
especially mass-produced wares like those of the Chimú Empire. Our work
with publicly-held collections poses little risk of increasing the
present commercial value of Chimú ceramics, while its public discussion
of the losses that accompany undocumented excavations and illicit
collection constitutes a benefit that outweighs any associated risks.

@@

Linear measurements from the sample of jars and bottles included vessel
height, rim thickness, and the diameter of each vessel at its orifice,
neck, belly, and base (Figure \ref{fig:linear}). Of these metrics,
vessel height, base diameter, and neck diameter did not differ
significantly between the two samples
(\href{https://seldenlab.github.io/peru.bottle.jar/linear-chim\%C3\%BA-vessel-metrics.html}{Supplementary
Materials, Ch. 1}). Overall, the jars and bottles in the sample overlap
in some key dimensions, with the bottles showing greater overall size
variations, which is consistent with production that uses a single set
of molds. The other three sets of linear measurements demonstrated
significant differences between the jars and bottles for orifice
diameter, belly diameter, and rim thickness
(\href{https://seldenlab.github.io/peru.bottle.jar/linear-chim\%C3\%BA-vessel-metrics.html}{Supplementary
Materials, Ch. 1}). Given the different production scenarios emerging
from archaeological field analyses---a combination of hand-shaping and
use of whole vessel or elemental molds, the significant differences in
orifice and belly diameters may indicate that the kinds of bottles and
jars in our sample were produced as distinct shapes, rather than by
\emph{mixing and matching} different molds to cater to changing consumer
tastes. The differences in rim thickness suggests that distinct
production practices were used when producing bottles and jars.

@@

Landmark data were aligned to a global coordinate system
\citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through
generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R
using the \texttt{geomorph} and \texttt{RRPP} packages
\citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition
translates and rotates the coordinate data to allow for comparisons
among objects, while also scaling each biface using unit-centroid
size---the square root of the sum of squared distances from each
landmark to the specimen's centroid
\citep{RN11139,RN11140,RN5622,RN478}. The \texttt{geomorph} package uses
a partial Procrustes superimposition that projects the aligned specimens
into tangent space subsequent to alignment in preparation for the use of
multivariate methods that assume linear space (Figure \ref{fig:gpa})
\citep{RN11141,RN11142,RN1646,RN11563}.

@@
